---
title: "Tidy Tuesday Exercise"
author: "Murphy John"
format: 
  html:
      toc: true
editor: visual
---

# Setup

```{r}
# load packages
library(dplyr)
library(ggplot2)
library(tidymodels)
library(MASS)
```

```{r}
# load data
tuesdata <- tidytuesdayR::tt_load('2025-04-08')
care_state <- tuesdata$care_state
```
Emergency room wait times vary significantly across the United States depending on factors such as hospital resources, patient volume, and staffing levels, with some states facing delays that can stretch for more than three hours.

Is there a connection between state populations and wait times?
Which conditions have the longest wait times? The shortest?

# Initial EDA
```{r}
summary(care_state)
str(care_state)
unique(care_state$condition)
unique(care_state$measure_id)
unique(care_state$measure_name)
```

# Research Question
What hospital resources and characteristics are associated with emergency department wait times?

# Data processing
```{r}
# consider only OP_18b as the outcome "Average (median) time patients spent in 
# the emergency department before leaving from the visit A lower number of 
# minutes is better"

dat_outcome <- care_state %>%
  filter(measure_id == "OP_18b") %>%
  dplyr::select(state, measure_id, score) %>%
  tidyr::pivot_wider(names_from = measure_id, values_from = score) %>%
  na.omit()
```

```{r}
# as predictors, consider HCP_COVID_19, IMM_3, OP_22, OP_23, OP_29, 
# SAFE_USE_OF_OPIOIDS, and SEP_1
dat_pred <- care_state %>%
  filter(measure_id %in% c("HCP_COVID_19", "IMM_3", "OP_22", "OP_23", "OP_29", 
                           "SAFE_USE_OF_OPIOIDS", "SEP_1")) %>%
  dplyr::select(state, measure_id, score) %>%
  tidyr::pivot_wider(names_from = measure_id, values_from = score)
```

```{r}
# merge outcome with predictors
dat_all <- dat_pred %>%
  right_join(dat_outcome, by="state")
```

# EDA
```{r, fig.height=5, fig.width=5}
# get the distribution of each variable
ggplot(dat_all, aes(x = OP_18b)) + 
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "OP_18b score") +
  theme_minimal()

# get mean and variance of outcome
mean(dat_all$OP_18b)
var(dat_all$OP_18b)

ggplot(dat_all, aes(x = HCP_COVID_19)) + 
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(title = "HCP_COVID_19 score") +
  theme_minimal()

ggplot(dat_all, aes(x = IMM_3)) + 
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(title = "IMM_3 score") +
  theme_minimal()

ggplot(dat_all, aes(x = OP_22)) + 
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "OP_22 score") +
  theme_minimal()

ggplot(dat_all, aes(x = OP_23)) + 
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(title = "OP_23 score") +
  theme_minimal()

ggplot(dat_all, aes(x = OP_29)) + 
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "OP_29 score") +
  theme_minimal()

ggplot(dat_all, aes(x = SAFE_USE_OF_OPIOIDS)) + 
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "SAFE_USE_OF_OPIOIDS score") +
  theme_minimal()

ggplot(dat_all, aes(x = SEP_1)) + 
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "SEP_1 score") +
  theme_minimal()
```

OP_18b is right-skewed, HCP_COVID_19 is right-skewed due to a few outliers, 
IMM_3 appears pretty normal aside from 3 outliers, OP_22 is right-skewed,
OP_23 is left-skewed and has a few outliers, OP_29 is normal with one outier,
SAFE_USE_OF_OPIOIDS is normal with one outlier, SEP_1 is pretty normal.

```{r}
# correlations
cor(dat_all[sapply(dat_all, is.numeric)])
```


# Model fits

```{r}
# split data into test and train
set.seed(333)
train_scheme <- sample(nrow(dat_all), 42)
dat_train <- dat_all[train_scheme, ]
dat_test <- dat_all[-train_scheme, ]
```

## Negative binomial regression
```{r}
# since we are dealing with count data that is over-dispersed, use negative 
# binomial regression

# set rec
nb1_recipe <- OP_18b ~ HCP_COVID_19 + IMM_3 + OP_22 + OP_23 + OP_29 + 
                    SAFE_USE_OF_OPIOIDS + SEP_1
# fit model
fit_nb1 <- MASS::glm.nb(nb1_recipe,
                        data = dat_train)

# summary
summary(fit_nb1)

# CIs
confint(fit_nb1)

## reduce the model to HCP_COVID_19, OP_22, and SEP_1
# set rec
nb2_recipe <- OP_18b ~ HCP_COVID_19 + OP_22

# fit model
fit_nb2 <- MASS::glm.nb(nb2_recipe,
                        data = dat_train)

# summary
summary(fit_nb2)

# CIs
confint(fit_nb2)

# get preds
preds_nb2 <- predict(fit_nb2, newdata = dat_train, type = "response")

# rmse
sqrt(mean((dat_train$OP_18b - preds_nb2)^2, na.rm = TRUE))
```

```{r}
# run nb model with cv- tidymodels doesnt support MASS objects so I'm improvising
set.seed(333)
nb_folds <- vfold_cv(dat_train, v = 5)

# function to train & evaluate nb model on each split
nb_cv_metrics <- function(split) {
  train_data <- analysis(split)
  test_data  <- assessment(split)
  
  # fit model
  model <- glm.nb(nb2_recipe, data = train_data)
  
  # get preds
  preds <- predict(model, newdata = test_data, type = "response")
  
  # get rmse
  rmse <- sqrt(mean((test_data$OP_18b - preds)^2, na.rm = TRUE))
  
  # get rsq
  log_likelihood_model <- logLik(model)
  log_likelihood_null <- logLik(glm.nb(OP_18b ~ 1, data = train_data))  # Null model (intercept only)
  r_squared <- 1 - (log_likelihood_model / log_likelihood_null)
  
  tibble(rmse = as.numeric(rmse), r_squared = as.numeric(r_squared))
}

# run
nb_metrics <- purrr::map_dfr(nb_folds$splits, nb_cv_metrics)

# results
mean(nb_metrics$rmse, na.rm = TRUE)
mean(nb_metrics$r_squared, na.rm = TRUE)
```

## Random forest
```{r}
set.seed(333)

# create folds
folds <- rsample::vfold_cv(dat_train, v = 5)

# specify and fit models
rf_spec <- 
  rand_forest() %>% 
  set_mode("regression") %>% 
  set_engine("ranger", seed = 333)

# wf
rf_wf <- workflows::workflow() %>%
  workflows::add_model(rf_spec) %>%
  workflows::add_formula(OP_18b ~ HCP_COVID_19 + OP_22)

fit_rf <- rf_wf %>%
  tune::fit_resamples(folds)

tune::collect_metrics(fit_rf)
```

## Natural Spline

```{r}
set.seed(333)

# spline spec
spline_spec <- 
  linear_reg() %>%
  set_engine(engine = 'lm') %>%
  set_mode('regression')

# natural spline recipe
spline_rec <- recipe(
  OP_18b ~ HCP_COVID_19 + OP_22,
  data=dat_train) %>%
  step_ns(HCP_COVID_19, deg_free = 3)

# workflow
spline_wf <- workflow() %>%
  add_recipe(spline_rec)

spline_fit <- spline_wf %>%
  add_model(spline_spec) %>%
  fit(data=dat_train)

# check variable selection
spline_fit %>% extract_fit_parsnip() %>% tidy()

# get pred
dat_train$pred_spline <- augment(spline_fit, new_data = dat_train)$.pred

# metrics
rmse(dat_train, truth=OP_18b, estimate = pred_spline)
rsq(dat_train, truth=OP_18b, estimate = pred_spline)
```

# Results

For each model, the following metrics are reported:

- Negative Binomial: RMSE = 33.02, R-Squared = 0.0506
- Random Forest: RMSE = 31.5, R-Squared = 0.257
- Natural Spline: RMSE = 24.8, R-Squared = 0.584

This suggests that the Natural spline model is performing the best. We will choose to further evaluate that model.

# Final model evaluation

```{r}
# get preds on test data
spline_preds <- augment(spline_fit, new_data = dat_test)

# residuals
spline_preds <- spline_preds %>%
  mutate(residual = OP_18b - .pred)

# RMSE
rmse(spline_preds, truth = OP_18b, estimate = .pred)

# R-squared
rsq(spline_preds, truth = OP_18b, estimate = .pred)
```

```{r}
# Residual plot
ggplot(spline_preds, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residuals vs. Predicted",
    x = "Predicted Values",
    y = "Residuals"
  ) + 
  scale_y_continuous(limits=c(-130, 130))
```

The natural spline model performs well on the test data with an RMSE = 44.7 and 
R-Squared = 0.715. The Residual vs Fitted plot shows in general a random pattern
with one outlying observation.

# Summary
The final model selected was a natural spline model considering "Percentage of 
healthcare personnel who are up to date with COVID-19 vaccinations" (HCP_COVID_19),
"Percentage of patients who left the emergency department before being seen 
Lower percentages are better" (OP_22) and "Percentage of patients who received 
appropriate care for severe sepsis and septic shock. Higher percentages are better"
(SEP_1) as predictors of "Average (median) time patients spent in the emergency 
department before leaving from the visit A lower number of minutes is better" 
(OP_18b). 

The model suggests that OP_22 is positively associated with wait times. For each 
unit increase in OP_22, the expected wait time increases by about 10.8 minutes.
The natural spline model revealed a nonlinear relationship between HCP_COVID_19 
and emergency department wait times. The three spline terms for HCP_COVID_19 
were all substantial in magnitude, with estimates of −39.3, 121.0, and 168.0 for 
the first, second, and third components respectively. These results suggest a 
non-monotonic association, where the effect of HCP_COVID_19 on wait times 
changes direction and intensity across its range. 