---
title: "fitting-exercise"
author: "Murphy John"
format: 
  html:
      toc: true
editor: visual
---

# Exercise 1

## Setup

```{r}
# load packages
library(ggplot2)
library(dplyr)
library(gtsummary)
```

```{r}
# load data
dat <- read.csv(
  here::here("fitting-exercise/data/Mavoglurant_A2121_nmpk.csv")
  )

names(dat) <- tolower(names(dat))
```

## Initial EDA

```{r, fig.width=10, fig.height=5}
# plot time by dv for each id, stratified by dose
plot <- ggplot(dat) +
  geom_line(aes(x=time, y=dv, group=id)) +
  facet_wrap(~dose)

print(plot)
```

## Initial processing

```{r}
# keep only obs where occ=1
dat1 <- dat %>%
  filter(occ == 1)
```

```{r}
# exclude obs where time=0, sum dv for each id
dat2 <- dat1 %>%
  filter(time != 0) %>%
  group_by(id) %>%
  summarize(
    Y = sum(dv)
  )

# create df with time=0 obs only
dat3 <- dat1 %>%
  filter(time == 0)

# join by id
dat4 <- full_join(x=dat2, y=dat3, by="id")
```

```{r}
# convert race and sex to factors
# keep only Y, dose, age, sex, race, wt, ht
dat5 <- dat4 %>%
  mutate(
    race = factor(race),
    sex = factor(sex)
  ) %>%
  select(
    Y, dose, age, sex, race, wt, ht
  )

str(dat5)

# save as rds
saveRDS(dat5, here::here("fitting-exercise/data/mavoglurant-processed.rds"))
```

## EDA

```{r}
# participant demographics by dose
table1 <- dat5 %>%
  gtsummary::tbl_summary(
    include = c(age, sex, race, wt, ht),
    by = dose
  ) %>%
  gtsummary::as_kable()

print(table1)
```

```{r, fig.width=3, fig.height=3}
# plot Y by dose
plot1 <- ggplot(dat5) +
  geom_boxplot(aes(x=factor(dose), y=Y))

print(plot1)
```

Boxplot of dose by Y should 2 individuals of dose 25 with higher Y values, 1 individual od dose 50 with higher Y value.

```{r, fig.width=3, fig.height=3}
# plot Y by age
plot2 <- ggplot(dat5) +
  geom_point(aes(x=age, y=Y))

print(plot2)
```

Scatterplot of age by Y appears randomly distributed.

```{r, fig.width=3, fig.height=3}
# plot Y by sex
plot3 <- ggplot(dat5) +
  geom_boxplot(aes(x=sex, y=Y))

print(plot3)
```

Boxplot of sex by Y shows one individual of sex 1 and one individual of sex 2 with higher values of Y.

```{r, fig.width=3, fig.height=3}
# distribution of Y
hist1 <- ggplot(dat5) +
  geom_histogram(aes(Y))

print(hist1)
```

Histogram of Y is right-skewed with some high-value outliers.

```{r, fig.width=3, fig.height=3}
# distribution of dose
bar1 <- ggplot(dat5) +
  geom_bar(aes(factor(dose)))

print(bar1)
```

Bar plot of dose shows that more individuals recieve dose 25 or 50 than dose 37.5.

```{r, fig.width=3, fig.height=3}
# distribution of age
hist2 <- ggplot(dat5) +
  geom_histogram(aes(age))

print(hist2)
```

Histogram of age appears multimodal. Most individuals are either late-20s or late-30s.

```{r, fig.width=3, fig.height=3}
# distribution of sex
bar2 <- ggplot(dat5) +
  geom_bar(aes(sex))

print(bar2)
```

Bar plot of sex shows that most individuals are of sex 1.

```{r, fig.width=3, fig.height=3}
# distribution of race
bar3 <- ggplot(dat5) +
  geom_bar(aes(race))

print(bar3)
```

Bar plot of race shows that most individuals are of race 1. Some are of race 2 and few are of races 7 and 88.

```{r, fig.width=3, fig.height=3}
# distribution of wt
hist3 <- ggplot(dat5) +
  geom_histogram(aes(wt))

print(hist3)
```

Histogram of weight shows a normal distribution.

```{r, fig.width=3, fig.height=3}
# distribution of ht
hist4 <- ggplot(dat5) +
  geom_histogram(aes(ht))

print(hist4)
```

Histogram of height shows normal to slightly left-skewed distribution.

```{r, fig.width=3, fig.height=3}
plot4 <- ggplot(dat5) +
  geom_boxplot(aes(x=sex, y=wt))

print(plot4)
```

Boxplot of sex by height shows that sex 1 has larger values for weight than sex 2

```{r, fig.width=3, fig.height=3}
plot5 <- ggplot(dat5) +
  geom_boxplot(aes(x=sex, y=ht))

print(plot5)
```

Boxplot of sex by height shows that sex 1 has larger values for height than sex 2

```{r, fig.height=10, fig.width=10}
# pair plot
pairs(dat5)
```

Pair plot shows general relationships between variables. Y seems positively associated with dose. Weight and height show a positive linear relationship.

## Model Fitting

### linear models

```{r}
# linear model to Y using dose
fit1 <- parsnip::linear_reg() %>%
  parsnip::fit(
    Y ~ dose, 
    data=dat5
  )

broom::tidy(fit1)

# predict Y
fit1_pred <- predict(
  fit1,
  new_data=dat5
)

# combine real data with predictions
fit1_result <- dat5 %>%
  select(Y, dose) %>%
  bind_cols(fit1_pred)

# calculate rmse
fit1_result %>%
  yardstick::rmse(
    truth=Y, 
    estimate=.pred
  )

# calculate R squared
fit1_result %>%
  yardstick::rsq(
    truth=Y, 
    estimate=.pred
  )
```

RMSE = 666 and R-squared = 0.516

```{r}
# linear model to Y using all predictors
fit2 <- parsnip::linear_reg() %>%
  parsnip::fit(
    Y ~ dose + age + sex + race + wt + ht, 
    data=dat5
  )

broom::tidy(fit2)

# predict Y
fit2_pred <- predict(
  fit2,
  new_data=dat5
)

# combine real data with predictions
fit2_result <- dat5 %>%
  select(Y, dose, age, sex, race, wt, ht) %>%
  bind_cols(fit2_pred)

# calculate rmse
fit2_result %>%
  yardstick::rmse(
    truth=Y, 
    estimate=.pred
  )

# calculate R squared
fit2_result %>%
  yardstick::rsq(
    truth=Y, 
    estimate=.pred
  )
```

RMSE = 591 and R-squared = 0.619

The full model, with all predictors included, has a lower RMSE, which indicated lower error, then the reduced model, with dose as the only predictor. The full model also has a higher R-squared value, indicating a better fit then the reduced model.

### logistic models

```{r}
# logistic model to sex using dose
fit3 <- parsnip::logistic_reg() %>%
  parsnip::fit(
    sex ~ dose, 
    data=dat5
  )

broom::tidy(fit3)

# predict sex
fit3_pred <- predict(
  fit3,
  new_data=dat5
)

# get prediction probabilities
fit3_prob <- predict(
  fit3,
  new_data = dat5,
  type = "prob"
)

# combine real data with predictions
fit3_result <- dat5 %>%
  select(sex, dose) %>%
  bind_cols(fit3_pred, fit3_prob)

# compute accuracy
yardstick::accuracy(
  fit3_result, 
  truth = sex,
  estimate = .pred_class
)

# compute ROC-AUC
yardstick::roc_auc(
  fit3_result, 
  truth=sex, 
  .pred_1
)
```

The model classification accuracy is about 86.7%.

The model has a ROC-AUC of 0.592 indicating predictive performance is low, just slightly better than random guessing.

```{r}
# logistic model to sex using all predictors
fit4 <- parsnip::logistic_reg() %>%
  parsnip::fit(
    sex ~ Y + dose + age + race + wt + ht, 
    data=dat5
  )

broom::tidy(fit4)

# predict sex
fit4_pred <- predict(
  fit4,
  new_data=dat5
)

# get prediction probabilities
fit4_prob <- predict(
  fit4,
  new_data = dat5,
  type = "prob"
)

# combine real data with predictions
fit4_result <- dat5 %>%
  select(sex, Y, dose, age, race, wt, ht) %>%
  bind_cols(fit4_pred, fit4_prob)

# compute accuracy
yardstick::accuracy(
  fit4_result, 
  truth = sex,
  estimate = .pred_class
)

# compute ROC-AUC
yardstick::roc_auc(
  fit4_result, 
  truth=sex, 
  .pred_1
)
```

The model classification accuracy is about 94.7%.

The model has a ROC-AUC of 0.980 indicating overall good predictive performance.

# Exercise 2

## Part 1

### Setup

```{r}
# set seed
rngseed = 1234
```

```{r}
# remove race variable
dat6 <- dat5 %>%
  select(!race)
```

```{r}
set.seed(rngseed)

# put 75% of the data into the training set 
data_split <- rsample::initial_split(dat6, prop = 3/4)

# Create data frames for the two sets:
train_data <- rsample::training(data_split)
test_data  <- rsample::testing(data_split)
```

### Model Fitting

#### Part 1

```{r}
# linear model to Y using dose
fit5 <- parsnip::linear_reg() %>%
  parsnip::fit(
    Y ~ dose, 
    data=train_data
  )

broom::tidy(fit5)

# predict Y
fit5_pred <- predict(
  fit5,
  new_data=train_data
)

# combine real data with predictions
fit5_result <- train_data %>%
  select(Y, dose) %>%
  bind_cols(fit5_pred)

# calculate rmse
fit5_result %>%
  yardstick::rmse(
    truth=Y, 
    estimate=.pred
  )

# calculate R squared
fit5_result %>%
  yardstick::rsq(
    truth=Y, 
    estimate=.pred
  )
```

RMSE = 703 and R-squared = 0.451

```{r}
# linear model to Y using all predictors
fit6 <- parsnip::linear_reg() %>%
  parsnip::fit(
    Y ~ dose + age + sex + wt + ht, 
    data=train_data
  )

broom::tidy(fit6)

# predict Y
fit6_pred <- predict(
  fit6,
  new_data=train_data
)

# combine real data with predictions
fit6_result <- train_data %>%
  select(Y, dose, age, sex, wt, ht) %>%
  bind_cols(fit6_pred)

# calculate rmse
fit6_result %>%
  yardstick::rmse(
    truth=Y, 
    estimate=.pred
  )

# calculate R squared
fit6_result %>%
  yardstick::rsq(
    truth=Y, 
    estimate=.pred
  )
```

RMSE = 627 and R-squared = 0.562

```{r}
# compute rmse of null model

null <- parsnip::null_model() %>%
  parsnip::set_engine("parsnip") %>%
  parsnip::set_mode("regression") %>%
  parsnip::fit(Y ~ ., data = train_data)

# predict Y
null_pred <- predict(
  null,
  new_data=train_data
)

# combine real data with predictions
null_result <- train_data %>%
  select(Y) %>%
  bind_cols(null_pred)

# calculate rmse
null_result %>%
  yardstick::rmse(
    truth=Y, 
    estimate=.pred
  )
```

RMSE = 948

##### Results

The RMSE of the full model is the lowest, at 627. This indicates that the full model is predicting values closer to the actual data than both the null and univariable models.

#### Part 2

```{r}
# setup

# fit model with resampling
set.seed(rngseed)

# define linear model specification
model_spec <- parsnip::linear_reg() %>% 
  parsnip::set_engine("lm") %>% 
  parsnip::set_mode("regression")

# define null model specification
null_spec <- parsnip::null_model() %>%
  parsnip::set_engine("parsnip") %>%
  parsnip::set_mode("regression")

# create folds
folds <- rsample::vfold_cv(train_data, v = 10)
```

```{r}
# linear model to Y using dose
fit_wf1 <- workflows::workflow() %>%
  workflows::add_model(model_spec) %>%
  workflows::add_formula(Y ~ dose)

fit_rs1 <- fit_wf1 %>%
  tune::fit_resamples(folds)

tune::collect_metrics(fit_rs1)
```

RMSE = 691, SE = 67.5; R-squared = 0.512, SE = 0.0592

```{r}
# linear model to Y using all predictors
fit_wf2 <- workflows::workflow() %>%
  workflows::add_model(model_spec) %>%
  workflows::add_formula(Y ~ dose + age + sex + wt + ht)

fit_rs2 <- fit_wf2 %>%
  tune::fit_resamples(folds)

tune::collect_metrics(fit_rs2)
```

RMSE = 646, SE = 64.8; R-squared = 0.573, SE = 0.0686

```{r}
# null model
fit_wf_null <- workflows::workflow() %>%
  workflows::add_model(null_spec) %>%
  workflows::add_formula(Y ~ .)

fit_rs_null <- fit_wf_null %>%
  tune::fit_resamples(folds)

tune::collect_metrics(fit_rs_null)
```

RMSE = 933, SE = 76.7

##### Results

The original RMSE value for the univariable model was 703, compared to the new value of 691. Using the cross-validation method to evaluate this model resulted in a better estimated fit. The standard error is 67.5, which is relatively large. This indicates that the RMSE varies considerably across folds, possibly suggesting an unstable model or high data variability.

The original RMSE value for the full model was 627, compared to the new value of 646. Using the cross-validation method to evaluate this model resulted in a weaker estimated fit. The standard error is 64.8, which is relatively large. This indicates that the RMSE varies considerably across folds, possibly suggesting an unstable model or high data variability.

The RMSE values for the null model are similar.

#### Changing the seed

```{r}
# setup

# fit model with resampling
set.seed(333)

# define linear model specification
model_spec <- parsnip::linear_reg() %>% 
  parsnip::set_engine("lm") %>% 
  parsnip::set_mode("regression")

# define null model specification
null_spec <- parsnip::null_model() %>%
  parsnip::set_engine("parsnip") %>%
  parsnip::set_mode("regression")

# create folds
folds <- rsample::vfold_cv(train_data, v = 10)
```

```{r}
# linear model to Y using dose
fit_wf1 <- workflows::workflow() %>%
  workflows::add_model(model_spec) %>%
  workflows::add_formula(Y ~ dose)

fit_rs1 <- fit_wf1 %>%
  tune::fit_resamples(folds)

tune::collect_metrics(fit_rs1)
```

RMSE = 706, SE = 48.1; R-squared = 0.484

```{r}
# linear model to Y using all predictors
fit_wf2 <- workflows::workflow() %>%
  workflows::add_model(model_spec) %>%
  workflows::add_formula(Y ~ dose + age + sex + wt + ht)

fit_rs2 <- fit_wf2 %>%
  tune::fit_resamples(folds)

tune::collect_metrics(fit_rs2)
```

RMSE = 645, SE = 52.7; R-squared = 0.566, SE = 0.0619

```{r}
# null model
fit_wf_null <- workflows::workflow() %>%
  workflows::add_model(null_spec) %>%
  workflows::add_formula(Y ~ .)

fit_rs_null <- fit_wf_null %>%
  tune::fit_resamples(folds)

tune::collect_metrics(fit_rs_null)
```

RMSE = 936, SE = 69.1

##### Results

Changing the seed results in different RMSE and SE values. For the univariable model, the new RMSE = 706, SE = 48.1; compared to the previous RMSE = 691, SE = 67.5. Note that if we had originally used this seed, we would have concluded that using cross-validation increases the RMSE for this model. For the full model, the new RMSe = 645, SE = 52.7; compared to the previous RMSE = 646, SE = 64.8. For the null model, the new RMSE = 936, SE = 69.1; compared to the previous RMSE = 933, SE = 76.7.

------------------------------------------------------------------------

## Part 2

### This section added by Shaun van den Hurk

### Load necessary libraries

```{r message=FALSE, warning=FALSE}
library(tidymodels)  
library(ggplot2)     
library(dplyr)       
library(tidyr) 
library(skimr)
```

We will put the training data and predicted vales from the original models (no CV) into a data frame.

```{r}
#Fit Models on Full Training Data (Not Using CV)

# Simple Model: Using dose only

lm_simple <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression") %>%
  fit(Y ~ dose, data = train_data)

# Full Model: Using All Predictors
lm_full <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression") %>%
  fit(Y ~ ., data = train_data)

# Null Model: Predicts the mean of Y for all observations
null_spec <- null_model() %>%
  set_engine("parsnip") %>%
  set_mode("regression")

null_fit <- null_spec %>%
  fit(Y ~ 1, data = train_data)

```

Using the models to extract the observed and predicted values

```{r}
#Extract Observed & Predicted Values for Each Model
#Get predictions for simple model
pred_simple <- predict(lm_simple, new_data = train_data) %>%
  bind_cols(train_data) %>%
  mutate(model = "Simple Model")

#Get predictions for full model
pred_full <- predict(lm_full, new_data = train_data) %>%
  bind_cols(train_data) %>%
  mutate(model = "Full Model")

#Get predictions for null model (mean prediction)
pred_null <- predict(null_fit, new_data = train_data) %>%
  bind_cols(train_data) %>%
  mutate(model = "Null Model")

#Combine Predictions into one dataframe
df_pred <- bind_rows(pred_simple, pred_full, pred_null) %>%
  select(Y, .pred, model)  # Keep observed values, predictions, and model label

#Pritn the dataframe
skim(df_pred)

#Create a plot for visualization (using ggplot)
ggplot(df_pred, aes(x = Y, y = .pred, color = model, shape = model)) +
  geom_point(alpha = 0.7, size = 2) +  # Scatter plot 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +  # 45-degree line
  scale_x_continuous(limits = c(0, 5000)) +
  scale_y_continuous(limits = c(0, 5000)) +
  labs(
    title = "Observed vs. Predicted Values for Three Models",
    x = "Observed Values",
    y = "Predicted Values",
    color = "Model",
    shape = "Model"
  ) +
  theme_minimal() 

```

The three horizontal lines in the Simple Model that only included the Dose varible is likely owing to the fact that Dose only takes three values in the dataset (25, 37.5 and 50) as these were likely the only different increments tested. Thus we see the grouping in the predictions cause by these three groups.

As stated, model 2 (the full model with all predictors) has the best performance and seems to sow some affinity towards the 45 degree line but not a close association and we see more values weighted to the lower left and not an even distribution.

We will make a plot of the predicted vs the residuals (predicted-observed) for Model 2. We will add a straight line at 0.

```{r}

# Compute residuals for the full model (Model 2)
residuals_full <- predict(lm_full, new_data = train_data) %>%
  bind_cols(train_data) %>%
  mutate(residuals = .pred - Y)  #Compute residuals

#Find maximum residual value to set symmetric y-axis limits
max_resid <- max(abs(residuals_full$residuals))  #Get max absolute residual

#Create Residual Plot
ggplot(residuals_full, aes(x = .pred, y = residuals)) +
  geom_point(alpha = 0.7, size = 2, color = "blue") +  #Scatter plot 
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +  #Reference line at 0
  scale_y_continuous(limits = c(-max_resid, max_resid)) +  #Symmetric y-axis limits
  labs(
    title = "Residual Plot for Full Model (Model 2)",
    x = "Predicted Values",
    y = "Residuals"
  ) +
  theme_minimal()  

```

We should see a general cloud of data with no real pattern for this. However that is not the case and we see some aggregation around the line (zero) and there are many negative numbers with higher values than the positive values.

### Model predictions and uncertainty

We are going to focus on model 2 going forward as it has better performance.

We are going to use a bootstrap method to help us sample the data and fit the model to the data. We start by reseting the seed to 1234 and create 100 bootstraps using the training data. We will then fit the model to these bootstrap samples and make predictions from the model for the original training data.

```{r}
#Reset random seed for reproducibility - 1234
set.seed(1234)

#Create 100 bootstrap samples from training data
boot_samples <- bootstraps(train_data, times = 100)  #Creates 100 bootstrap datasets from training data

#Define a function to fit Model 2 and make predictions and to extract the training data from the bootstrap sample
fit_and_predict <- function(split) {boot_train <- analysis(split)  #Extract training set from bootstrap sample
  lm_boot <- linear_reg() %>%   #Fit the Full Model to the bootstrap training set
    set_engine("lm") %>%
    set_mode("regression") %>%
    fit(Y ~ ., data = boot_train)
  predict(lm_boot, new_data = train_data) %>% #Predict on original training data
    pull(.pred)}  #Extract predicted values as a numeric vector


boot_preds <- map(boot_samples$splits, fit_and_predict)  #Apply model to each bootstrap sample & store predictions

#Convert list of predictions to data frame
boot_pred_df <- as.data.frame(do.call(cbind, boot_preds))
colnames(boot_pred_df) <- paste0("boot_", seq(1, length(boot_preds)))  #Label columns

#Add original observed values for reference
boot_pred_df <- bind_cols(train_data %>% select(Y), boot_pred_df)

#Print a preview of the bootstrap predictions
print("First few rows of bootstrap predictions:")
print(head(boot_pred_df))

```

Above we see a preview of the bootstrap predictions that we modeled.

We will now calculate the mean, median and confidence intervals for the bootstrap predictions and generate a dataframe with this and provide a summary/overview of this dataframe of these values combined with Y (observed vaues).

```{r}
#Convert list of predictions to a matrix (for easier calculations)
pred_matrix <- do.call(cbind, boot_preds)  #convert list to a matrix
colnames(pred_matrix) <- paste0("boot_", seq(1, ncol(pred_matrix)))  #label columns

#Compute Mean, Median & 89% Confidence Intervals
preds <- pred_matrix %>% apply(1, function(x) {c(Mean_Prediction = mean(x), quantile(x, c(0.055, 0.5, 0.945)))}) %>%
  t()  #Transpose to align columns correctly

#Convert to dataframe and add `Y` 
pred_summary <- as.data.frame(preds) %>% rename(Mean_Prediction = `Mean_Prediction`,
    CI_Lower_89 = `5.5%`,
    Median_Prediction = `50%`,
    CI_Upper_89 = `94.5%`) %>%
  mutate(Y = train_data$Y)  

#We will verify this step because differences in size will cause failure in the following steps
#Verify data structure to ensure correct fix
print(paste("Size of train_data:", nrow(train_data)))  #Should be 90
print(paste("Size of pred_summary:", nrow(pred_summary)))  #Should also be 90

print("First few rows of bootstrap prediction summary with Y:")
print(head(pred_summary))
str(pred_summary)  

```

We will now geenrate a plot to visualise the data. We want to evaluate the observed values, the point estimates and the calcuated statistics from the bootstrap predictions. We plot the Observed vs Predicted values wirh bootstrap confidence intervals, mean and median statistics.

```{r}

#Get Original point estimate from the full model (Model 2)
point_estimates <- predict(lm_full, new_data = train_data) %>%
  bind_cols(train_data) %>% rename(Point_Estimate = .pred)  #Rename prediction column

#Use left_join() to ensure correct data alignment matched based on Y
pred_plot_data <- point_estimates %>% left_join(pred_summary, by = "Y") %>% select(Y, Point_Estimate, Median_Prediction, CI_Lower_89, CI_Upper_89)  

#Create plot for visualization
ggplot(pred_plot_data) +
  #black dots for original point estimates
  geom_point(aes(x = Y, y = Point_Estimate), color = "black", size = 2, alpha = 0.8) +
    # Blue dots for bootstrap median predictions
  geom_point(aes(x = Y, y = Median_Prediction), color = "blue", size = 2, alpha = 0.7) +
    # Red dots for lower bound of 89% confidence interval
  geom_point(aes(x = Y, y = CI_Lower_89), color = "red", size = 2, alpha = 0.5) +
    # Green dots for upper bound of 89% confidence interval
  geom_point(aes(x = Y, y = CI_Upper_89), color = "green", size = 2, alpha = 0.5) +
    # Add a 45-degree line (ideal case where predicted = observed)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
    # Ensure x and y axes match for comparison
  scale_x_continuous(limits = c(0, 5000)) +
  scale_y_continuous(limits = c(0, 5000)) +
    # Labels and theme
  labs(title = "Observed vs. Predicted Values with Bootstrap Confidence Intervals",
    x = "Observed Values", y = "Predicted Values", color = "Prediction Type") + theme_minimal() 

```

It is hard to distinguish all of the different points without a legend and to see some of the colours. So the same plot will be generated using different symbols and providing a legend.

```{r}
#Get original point estimate from the full model (Model 2)
point_estimates <- predict(lm_full, new_data = train_data) %>%
  bind_cols(train_data) %>% rename(Point_Estimate = .pred)  #Rename prediction column

#Ensure Correct data alignment Using left_join()
pred_plot_data <- point_estimates %>% left_join(pred_summary, by = "Y") %>% select(
    Y,                  #Observed values 
    Point_Estimate,     #Original model predictions
    Mean_Prediction,    #Mean bootstrap prediction
    Median_Prediction,  #Median bootstrap prediction
    CI_Lower_89,        #Lower 89% confidence interval
    CI_Upper_89)         #Upper 89% confidence interval

ggplot(pred_plot_data, aes(x = Y)) +
  # Black dots for original point estimates (Model 2 predictions)
  geom_point(aes(y = Point_Estimate, color = "Point Estimate", shape = "Point Estimate"), size = 2, alpha = 0.8) +
    # Purple dots for mean bootstrap prediction
  geom_point(aes(y = Mean_Prediction, color = "Mean Prediction", shape = "Mean Prediction"), size = 2, alpha = 0.7) +
  # Blue dots for bootstrap median predictions
  geom_point(aes(y = Median_Prediction, color = "Median Prediction", shape = "Median Prediction"), size = 2, alpha = 0.7) +
  # Red dots for lower bound of 89% confidence interval
  geom_point(aes(y = CI_Lower_89, color = "CI Lower Bound", shape = "CI Lower Bound"), size = 2, alpha = 0.5) +
  # Green dots for upper bound of 89% confidence interval
  geom_point(aes(y = CI_Upper_89, color = "CI Upper Bound", shape = "CI Upper Bound"), size = 2, alpha = 0.5) +
  # Add a 45-degree line (ideal case where predicted = observed)
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  scale_x_continuous(limits = c(0, 5000)) +
  scale_y_continuous(limits = c(0, 5000)) +
  # Manually define colors and shapes with proper legend
  scale_color_manual(name = "Prediction Type", values = c(
      "Point Estimate" = "black",
      "Mean Prediction" = "purple",
      "Median Prediction" = "blue",
      "CI Lower Bound" = "red",
      "CI Upper Bound" = "green")) +
  scale_shape_manual(name = "Prediction Type", values = c(
      "Point Estimate" = 16,
      "Mean Prediction" = 17,
      "Median Prediction" = 18,
      "CI Lower Bound" = 3,
      "CI Upper Bound" = 3 )) +
  # Labels and theme
  labs(title = "Observed vs. Predicted Values with Bootstrap Confidence Intervals", x = "Observed Values (Y)",
    y = "Predicted Values") + theme_minimal()

```

Model 2 seems to capture the trend of observed vs predicted (the 45% line) quite well although not perfectly. The values from the bootstrap sampling median and mean demonstrate a similar pattern and so that shows the bootstrap sampling did not significantly shift the trend. The confidence intervals also follow this general tredn and are quite close which might indicate that the models predictions are quite stable in many cases (although the gap increases in some areas).

------------------------------------------------------------------------

## Part 3

### Final evaluation using test data

```{r}
# use model 2 to predict for test data
pred_test <- predict(
  fit6,
  new_data=test_data
)

# combine real data with predictions
test_result <- test_data %>%
  select(Y, dose, age, sex, wt, ht) %>%
  bind_cols(pred_test)
```

```{r}
# plot predicted vs observed for both training data and test data
ggplot() +
  geom_point(data=fit6_result, aes(x=Y, y=.pred, color="Train Data")) +
  geom_point(data=test_result, aes(x=Y, y=.pred, color="Test Data")) +
  geom_abline(slope = 1, intercept = 0) +
  scale_x_continuous(limits=c(0,5700)) +
  scale_y_continuous(limits=c(0,5700)) +
  scale_color_manual(values = c("Train Data" = "black", "Test Data" = "coral3"), name="Dataset")
```

#### Results

Based on the metrics, the multivariable model performs better than the null model. The RMSE of the full model is 627, with an R-squared of 0.562, which is an improvement over the null model’s RMSE of 948. A smaller RMSE indicates that the full model predicts values closer to the actual data than the null model. The R-squared value means that the full model explains 56.2% of the variation in the data beyond what the null model accounts for.

The univariable model with dose also performs better than the null model, with an RMSE of 706 and an R-squared value of 0.451. Compared to the null model, the smaller RMSE suggests that the univariable model predicts values closer to the actual data. The R-squared value indicates that the univariable model explains 45.1% of the variation in the data beyond the null model. The univariable model is useful for quantifying and visually representing the effect of dose on the outcome. However, beyond this, its practicality may be limited, as the full model demonstrates that in real data, factors beyond dose influence the outcome.

Comparing the metrics of the multivariable and univariable models shows that the multivariable model has a smaller RMSE and a larger R-squared. This suggests that the full model predicts values more accurately than the univariable model and explains more of the variation in the data beyond the null model. This result aligns with the general expectation that as the number of predictors in a model increases, so does the model’s fit.

Importantly, using train and test datasets reveals that the full model does not appear to be overfitting the training data. This is a crucial feature to assess when evaluating the practicality of a model. If one aims to determine an appropriate drug dose for an individual based on their characteristics, the full model could be useful in practice.
