---
title: "Synthetic Data Generation and Exploration"
author: "Murphy John"
editor: visual
---

# Introduction

For this exercise, I am going to simulate simple data from a (hypothetical) vaccine cohort of individuals who received various doses of a vaccine. Samples were collected and a specific antibody was measured on days 0, 3, 5, 7, 14, 21, 28, and 35. The following variables are to be included: `id`, `age`, `dose`, `days`, `titer.` Brief descriptions of the variables are as follows:

-   `id` is an identifier for each participant
-   `age` is the participant's age at day 0
-   `dose` is the numerical dose of vaccine for which the individual received in micrograms (mcg) (10, 100, 1000)
-   `days` represents the number of days since vaccination on which the sample was collected
-   `titer` is the measured amount of antibody in the sample

The simulation assumes the following:

-   The immune response follows a logistic growth function up to a peak, followed by an exponential decay.
-   Older individuals have a weaker boost in response to vaccination.
-   Baseline titers (day 0) are independent of dose and vary between individuals.
-   Individual variation is present in peak titer, growth rate, decay rate, and peak timing.
-   Titers are strictly positive.

# Setup

## General

```{r}
# load packages
library(dplyr)
library(ggplot2)
```

## Assign parameters

```{r}
# Set seed for reproducibility
set.seed(123)

# number of individuals
N <- 100

# assign random ages
ages <- sample(18:80, N, replace = TRUE)

# vaccine doses
dose <- c(10, 100, 1000)

# measurement times
days <- c(0, 3, 5, 7, 14, 21, 28, 35) 

# assign doses
doses <- rep(dose, length.out = N)

# categorical dose variable
dose_cat <- if_else(
  doses == 10, "low", 
  if_else(
    doses == 100, "medium", "high"
  )
)
```

# Simulate data

```{r}
# generate baseline titers
baseline_titers <- runif(N, min = 0, max = 50)

# function to simulate titer with logistic growth and exponential decay
simulate_titer <- function(age, dose, days, baseline) {
  boost_factor <- exp(-0.05 * (age - 18))
  peak_titer <- dose * boost_factor * 1000 * rnorm(1, mean = 1, sd = 0.2) 
  growth_rate <- 0.3 * rnorm(1, mean = 1, sd = 0.01)
  decay_rate <- 0.2 * rnorm(1, mean = 1, sd = 0.01)  
  peak_day <- 7 + rnorm(1, mean = 0, sd = 3) 
  
  # logistic growth until peak day, then exponential decay
  titer <- ifelse(days == 0, baseline,
                  ifelse(days <= peak_day,
                         baseline + (peak_titer - baseline) / (1 + exp(-growth_rate * (days - peak_day / 2))),
                         pmax(peak_titer * exp(-decay_rate * (days - peak_day)), 0)))
  
  return(titer)
}

# Generate dataset
vaccine_data <- expand.grid(ID = 1:N, days = days) %>%
  mutate(
    age = ages[ID],  
    dose = doses[ID],  
    dose_cat = dose_cat[ID],
    baseline_titer = baseline_titers[ID],
    titer = mapply(simulate_titer, age, dose, days, baseline_titer),
    log_titer = log(titer)
  )
```

# Data exploration

```{r}
# view summary
summary(vaccine_data)

# plot days by titer stratified by dose
ggplot(vaccine_data) +
  geom_line(aes(x = days, y = log_titer, color = dose_cat, group = ID)) +
  geom_point(aes(x = days, y = log_titer, color = dose_cat)) +
  labs(title = "log(Titer) over time by Dose", 
       x = "Day", y = "log(Titer)", color = "Dose") +
  theme_minimal()

# calculate the mean titer for each dose and day combination
mean_titer_data <- vaccine_data %>%
  group_by(dose, days) %>%
  summarise(mean_log_titer = mean(log_titer), .groups = "drop")

# plot the individual points and mean curves by dose
ggplot(vaccine_data) +
  geom_point(aes(x = days, y = log_titer, color = factor(dose)), alpha = 0.5) +
  geom_line(data = mean_titer_data, aes(x = days, y = mean_log_titer, color = factor(dose)), size = 1.2) + 
  labs(title = "Mean log(Titer) Over time by Dose", 
       x = "Day", y = "log(Titer)", color = "Dose") +
  theme_minimal()
```

Wile some overlap is apparent, dose seems to be a distinguishing factor of titer. This is good considering that a positive association between dose and titer was assummed. We also assumed that age was negatively associated with titer. Let's explore this in the data.

```{r}
# create age categories
vaccine_data2 <- vaccine_data %>%
  mutate(age_group = case_when(
    age < 30 ~ "18-29",
    age < 50 ~ "30-49",
    age < 70 ~ "50-69",
    TRUE ~ "70+"
  ))

mean_titer_age <- vaccine_data2 %>%
  group_by(dose, days, age_group) %>%
  summarise(mean_log_titer = mean(log_titer), .groups = "drop")

# plot low dose by age group
ggplot(subset(vaccine_data2, dose == 10)) +  # Filter data for low dose
  geom_point(aes(x = days, y = log_titer, color = age_group), alpha = 0.5) +
  geom_line(data = subset(mean_titer_age, dose == 10), 
            aes(x = days, y = mean_log_titer, color = age_group, group = age_group), size = 1.2) +
  labs(title = "Mean log(Titer) Over Time for Low Dose by Age Group", 
       x = "Day", y = "log(Titer)", color = "Age Group") +
  theme_minimal()

# plot medium dose by age group
ggplot(subset(vaccine_data2, dose == 100)) +  # Filter data for low dose
  geom_point(aes(x = days, y = log_titer, color = age_group), alpha = 0.5) +
  geom_line(data = subset(mean_titer_age, dose == 100), 
            aes(x = days, y = mean_log_titer, color = age_group, group = age_group), size = 1.2) +
  labs(title = "Mean log(Titer) Over Time for Medium Dose by Age Group", 
       x = "Day", y = "log(Titer)", color = "Age Group") +
  theme_minimal()

# plot high dose by age group
ggplot(subset(vaccine_data2, dose == 1000)) +  # Filter data for low dose
  geom_point(aes(x = days, y = log_titer, color = age_group), alpha = 0.5) +
  geom_line(data = subset(mean_titer_age, dose == 1000), 
            aes(x = days, y = mean_log_titer, color = age_group, group = age_group), size = 1.2) +
  labs(title = "Mean log(Titer) Over Time for High Dose by Age Group", 
       x = "Day", y = "log(Titer)", color = "Age Group") +
  theme_minimal()
```

The age differences that we specified when simulating the data are certainly present in our final dataset.

This simulation was inspired by real data but certainly simplified for the purpose of this exercise. In the future, one could consider additional covariates, such as repeat vaccination status, prior infections, and additional comorbidities, all of which are known to be associated with immune response to vaccination.



