---
title: "Data Analysis Exercise"
subtitle: "cdcdata-exercise"
author: "Murphy John"
format: html
editor: 
  markdown: 
    wrap: 72
---

Data Description: Data was downloaded from the CDC on 2025-02-03. This
dataset is titled NHANES Select Chronic Conditions Prevalence Estimates.
It contains national data on obesity, high total cholesterol, and
hypertension rates from 1999 to 2018 by sex, race, and age.There are a
total of 950 rows and 12 columns. See
[here](https://data.cdc.gov/NCHS/NHANES-Select-Chronic-Conditions-Prevalence-Estima/i667-sjhg/about_data)
for more details.

# Setup

```{r}
# load packages
library(dplyr)
library(ggplot2)
```

```{r}
# loads data
dat <- read.csv(here::here("cdcdata-exercise/data/raw/rawdata.csv"))
```

# Data processing

```{r}
# view data summary
summary(dat)

# select a subset of variables
dat1 <- dat %>%
  select(
    years = Survey.Years,
    sex = Sex,
    age = Age.Group,
    measure = Measure,
    percent = Percent
  ) 
```

```{r}
# filter for more granular data - separate by sex, age categories
dat2 <- dat1 %>%
  filter(
    sex != "All",
    age != "20 and over",
    age != "2-19"
  ) %>%
  mutate(
    age_factor = factor(age, levels = c(
      "2-5","6-11","12-19","20-29","30-39","40-49",
      "50-59","60-69","70 and over")
    ),
    age = as.numeric(age_factor),
    sex = factor(sex, levels = c("Female", "Male")),
    years_factor = factor(years, levels = c(
      "1999-2000","2001-2002","2003-2004",
      "2005-2006","2007-2008","2009-2010",
      "2011-2012","2013-2014","2015-2016","2017-2018"
    )),
    years = as.numeric(years_factor)
  )

# turn obesity, cholesterol, and hypertension into their own variables
dat3 <- dat2 %>%
  group_by(years, sex, age) %>%
  tidyr::pivot_wider(
    names_from = measure,
    values_from = percent
  )
```

# Exploratory data analysis

```{r}
# number of rows
nrow(dat3)

# number of columns
ncol(dat3)

summary(dat3)

# print first few obs
head(dat3)

# print last few obs
tail(dat3)
```

It looks like there is some missing data. Locate those missing
observations.

```{r}
# count NAs by column
colSums(is.na(dat3))

# filter data for only rows with NA values
dat_na <- dat3 %>% filter(if_any(everything(), is.na))
```

`High Total Cholesterol` and `Hypertension` data was not collected for
those aged less than 19 years. Remove these observations to focus on the
relationship between the three measures.

# Briefly, data processing

```{r}
dat4 <- dat3 %>% 
  dplyr::rename(
    obesity = Obesity,
    hypertension = Hypertension,
    cholesterol = `High Total Cholesterol`
  ) %>%
  na.omit()

summary(dat4)
```

# More EDA

Explore each variable's categories and/or distributions.

```{r}
# categorical variables
table(dat4$years, dat4$sex)
table(dat4$years, dat4$age)
table(dat4$sex, dat4$age)

# numeric variables
summary(dat4$obesity)
ggplot(dat4, aes(x = obesity)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") 

summary(dat4$cholesterol)
ggplot(dat4, aes(x = cholesterol)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") 

summary(dat4$hypertension)
ggplot(dat4, aes(x = hypertension)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") 
```

Lets look at the relationship between years and our three measurements.

```{r}
# obesity
plot(dat4$years, dat4$obesity)
summary(lm(obesity ~ years, data=dat4))

# hypertension
plot(dat4$years, dat4$hypertension)
summary(lm(hypertension ~ years, data=dat4))

# cholesterol
plot(dat4$years, dat4$cholesterol)
summary(lm(cholesterol ~ years, data=dat4))
```

For every one unit increase in year, there is a $1.2$ unit increase in
obesity. For every one unit increase in year, there is a $0.9$ unit
decrease in cholesterol.

Lets look at the relationship between sex and our three measurements.

```{r}
# obesity
plot(dat4$sex, dat4$obesity)
summary(lm(obesity ~ sex, data=dat4))

# hypertension
plot(dat4$sex, dat4$hypertension)
summary(lm(hypertension ~ sex, data=dat4))

# cholesterol
plot(dat4$sex, dat4$cholesterol)
summary(lm(cholesterol ~ sex, data=dat4))
```

Male sex has $3.03$ unit decrease in obesity from female sex. Male sex
has $3.45$ unit decrease in cholesterol from female sex.

Lets look at the relationship between age and our three measurements.

```{r}
# obesity
plot(dat4$age, dat4$obesity)
dat_age1 <- dat4 %>%
  filter(
    age != 9
  )
summary(lm(obesity ~ age, data=dat_age1))

dat_age2 <- dat4 %>%
  filter(
    age %in% c(4, 9)
  )
summary(lm(obesity ~ age, data=dat_age2))


# hypertension
plot(dat4$age, dat4$hypertension)
summary(lm(hypertension ~ age, data=dat4))

# cholesterol
plot(dat4$age, dat4$cholesterol)
summary(lm(cholesterol ~ age, data=dat4))
```

Excluding the final age category, for every subsequent age group, there
is a $3.14$ unit increase in obesity. There is no difference in obesity
between groups 4 and 9. For every subsequent age group, there is a
$13.56$ unit increase in hypertension. For every subsequent age group,
there is a $1.64$ unit increase in cholesterol.

# EDA Conclusions

There are 10 2-year `years` categories from 1999-2018. For each `years`
category, there is an observation for each `sex` (Female, Male), and
within each sex, an observation for each of 6 age groups
(20-29,30-39,40-49,50-59,60-69,70 and over). This makes $6*2=12$
observations each year and $10*12=120$ total observations in the data.
Each observation has a value for percentage of population with
`obesity`, high `cholesterol`, and `hypertension`. 

Obesity overall has
mean 35.31 with (30.57, 40.90) quartiles. For every one unit increase in
year group, there is a $1.2$ unit increase in obesity. Male sex has
$3.03$ unit decrease in obesity from female sex. Excluding the final age
category, for every subsequent age group, there is a $3.14$ unit
increase in obesity. There is no difference in obesity between age
categories "20-29" and "70 and over".

Hypertension overall has mean 49.98 with (30.65, 70.90) quartiles. There
is no obvious relationship between hypertension and year or sex. For
every subsequent age group, there is a $13.56$ unit increase in
hypertension.

Cholesterol overall has mean 14.35 with (8.5, 18.55) quartiles. For
every one unit increase in year, there is a $0.9$ unit decrease in
cholesterol. Male sex has $3.45$ unit decrease in cholesterol from
female sex. For every subsequent age group, there is a $1.64$ unit
increase in cholesterol.

# END OF PART 1

# This section contributed by Asmith Joseph 

# Running the Data 
To read the data, I set my working directory to match the project structure using setwd("C:/Users/ajose35/Desktop/Ecole/MADA-course/murphyjohn-MADA-portfolio/coding-exercise"). Since the dataset was located in a different folder than expected, I used here::i_am("cdcdata-exercise/data/raw/rawdata.csv") to manually define the project root without modifying my collaborator’s code. After ensuring that here correctly resolved the file path by running file.exists(here::here("cdcdata-exercise/data/raw/rawdata.csv")), I successfully loaded the dataset using dat <- read.csv(here::here("cdcdata-exercise/data/raw/rawdata.csv")). This approach allowed me to work with the existing project structure while maintaining compatibility with my collaborator’s code.

# Generate Synthetic Data with the Same Structure

Based on the summary the dataset  contains:

Categorical variables: Survey.Years, Sex, Age.Group, Measure
Numeric variables: Percent, Standard.Error, Lower.95..CI.Limit, Upper.95..CI.Limit. The create the Synthetic data I used this prompt "Generate R code to create synthetic data with 950 observations, including categorical and numeric variables that resemble the original dataset structure." 

```{r}
# Loading necessary libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(fabricatr)  # For realistic synthetic data

# Set seed for reproducibility
set.seed(123)

# Creating synthetic dataset
synthetic_data <- tibble(
  Survey.Years = sample(c("1999-2000", "2001-2002", "2003-2004",
                          "2005-2006", "2007-2008", "2009-2010",
                          "2011-2012", "2013-2014", "2015-2016", "2017-2018"), 
                        950, replace = TRUE),
  Sex = sample(c("Male", "Female"), 950, replace = TRUE),
  Age.Group = sample(c("2-5", "6-11", "12-19", "20-29", "30-39", "40-49",
                       "50-59", "60-69", "70 and over"), 950, replace = TRUE),
  Measure = sample(c("Obesity", "High Total Cholesterol", "Hypertension"), 950, replace = TRUE),
  Percent = round(runif(950, min = 3, max = 90), 1),  # Simulated percentages
  Standard.Error = round(runif(950, min = 0.5, max = 7.5), 2),
  Lower.95..CI.Limit = Percent - runif(950, min = 1.5, max = 8.0),
  Upper.95..CI.Limit = Percent + runif(950, min = 1.5, max = 8.0)
)

# Saving synthetic data
write.csv(synthetic_data, "synthetic_data.csv", row.names = FALSE)

# View first few rows
head(synthetic_data)

```

# Cleaning  and Processing the Synthetic Data

In the cleaning process I performed data selection, filtering, and transformation.

```{r}
# Cleaning  and Processing the Synthetic Data

# Selecting relevant variables
synthetic_dat1 <- synthetic_data %>%
  select(
    years = Survey.Years,
    sex = Sex,
    age = Age.Group,
    measure = Measure,
    percent = Percent
  )

# Filtering granular data - exclude "All" and broad age categories
synthetic_dat2 <- synthetic_dat1 %>%
  filter(
    sex != "All",
    age != "20 and over",
    age != "2-19"
  ) %>%
  mutate(
    age_factor = factor(age, levels = c(
      "2-5", "6-11", "12-19", "20-29", "30-39", "40-49",
      "50-59", "60-69", "70 and over")
    ),
    age = as.numeric(age_factor),
    sex = factor(sex, levels = c("Female", "Male")),
    years_factor = factor(years, levels = c(
      "1999-2000", "2001-2002", "2003-2004",
      "2005-2006", "2007-2008", "2009-2010",
      "2011-2012", "2013-2014", "2015-2016", "2017-2018"
    )),
    years = as.numeric(years_factor)
  )

# CHECK FOR DUPLICATES BEFORE PIVOTING
synthetic_dat2 %>%
  group_by(years, sex, age, measure) %>%
  summarise(n = n(), .groups = "drop") %>%
  filter(n > 1)

# Summarizing percent values to ensure uniqueness before pivoting
synthetic_dat3 <- synthetic_dat2 %>%
  group_by(years, sex, age, measure) %>%
  summarise(percent = mean(percent, na.rm = TRUE), .groups = "drop") %>%  # Take mean to resolve duplicates
  tidyr::pivot_wider(
    names_from = measure,
    values_from = percent
  )

# Rename columns to match cleaned dataset
synthetic_dat4 <- synthetic_dat3 %>%
  rename(
    obesity = Obesity,
    hypertension = Hypertension,
    cholesterol = `High Total Cholesterol`
  ) %>%
  na.omit()

```

# Summarizing and Exploring the Synthetic Data

```{r}
summary(synthetic_dat4)


colSums(is.na(synthetic_dat4))

```

```{r}
sum(is.na(synthetic_dat4$obesity))  # Check number of NA values in obesity

```

# Visualizion 

```{r}
#Histogram of Obesity Rates


ggplot(synthetic_dat4 %>% filter(!is.na(obesity)), aes(x = obesity)) +
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.6, color = "black") +
  theme_minimal() +
  labs(title = "Obesity Rate Distribution ", x = "Obesity Rate", y = "Count")


```

```{r}
#Boxplot of Cholesterol by Sex

ggplot(synthetic_dat4, aes(x = sex, y = cholesterol, fill = sex)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Cholesterol Levels by Sex", x = "Sex", y = "Cholesterol Level")

```

```{r}
# Line Plot of Hypertension Over Years
ggplot(synthetic_dat4, aes(x = years, y = hypertension, color = sex)) +
  geom_point(alpha = 0.5) +  # Add scatter points for better visibility
  geom_smooth(method = "loess", se = FALSE, size = 1) +  # Add smoothed trend line
  theme_minimal() +
  labs(title = "Hypertension Levels Over Years", 
       x = "Year", y = "Hypertension Rate") +
  scale_x_continuous(breaks = unique(synthetic_dat4$years))  # Ensure year labels are readable


```

```{r}
# Bar Chart Hypertension Levels by Year
ggplot(synthetic_dat4, aes(x = as.factor(years), y = hypertension, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") +  
  theme_minimal() +
  labs(title = "Hypertension Levels by Year", 
       x = "Year", y = "Hypertension Rate") +
  scale_fill_manual(values = c("blue", "pink"))  # Custom colors for Male/Female

```

# Regression Models
Running Regression Models on the Synthetic Data Because these models will help analyze trends, test hypotheses, and validate data structure. Addiionally, they will assess associations between health indicators and predictors, checking for statistical significance and identifying gaps where more variables may improve predictions.

```{r}
# Looking at Obesity Trends Over Years
summary(lm(obesity ~ years, data = synthetic_dat4))


# Looking at Hypertension by Sex
summary(lm(hypertension ~ sex, data = synthetic_dat4))



# Looking at Cholesterol by Age
summary(lm(cholesterol ~ age, data = synthetic_dat4))

```

# Interpretation of Regression Outputs
Model	                  Predictor   	       Effect	        Significance (p-value)	  Interpretation
Obesity ~ Years	        Years	              -0.488 (↓)	    0.447	                    No significant trend in obesity over time
Hypertension ~ Sex    	Male vs. Female   	+1.711 (↑)	    0.653	                    No significant difference in hypertension between sexes
Cholesterol ~ Age	      Age	                +0.164 (↑)	    0.811	                    No significant impact of age on cholesterol

The regression analysis shows no significant relationships between obesity, hypertension, or cholesterol and their predictors. Obesity trends over time show a slight decline (-0.488 per year), but this effect is not significant (p = 0.447). Hypertension rates are slightly higher in males (+1.711), but the difference is insignificant (p = 0.653). Cholesterol levels increase slightly with age (+0.164), but age has no meaningful impact (p = 0.811).

All models have low R-squared values, indicating weak predictive power. The dataset likely lacks key variables or requires better transformations to capture meaningful patterns. 

#  Comparing the Synthetic and Original Data
This comparison will allow me to check similarity, compare means and standard deviations.

```{r}
#  Comparing the Synthetic and Original Data

comparison_table <- tibble(
  Variable = c("Obesity", "Cholesterol", "Hypertension"),
  Original_Mean = c(35.31, 14.35, 49.98),  # Replace with real means
  Synthetic_Mean = c(mean(synthetic_dat4$obesity), mean(synthetic_dat4$cholesterol), mean(synthetic_dat4$hypertension)),
  Original_SD = c(6.58, 6.94, 24.05),
  Synthetic_SD = c(sd(synthetic_dat4$obesity), sd(synthetic_dat4$cholesterol), sd(synthetic_dat4$hypertension))
)

print(comparison_table)

```

The comparison between the synthetic and original datasets shows some differences in means and standard deviations for obesity, cholesterol, and hypertension. Obesity levels in the synthetic data (Mean = 46.36, SD = 17.97) are higher than the original (Mean = 35.31, SD = 6.58), suggesting a wider distribution. Cholesterol levels also appear inflated in the synthetic dataset (Mean = 45.90 vs. 14.35), with a higher standard deviation (17.67 vs. 6.94), indicating greater variability. Hypertension levels, however, remain closer to the original data (Mean = 48.25 vs. 49.98, SD = 19.10 vs. 24.05), showing a better match.

Overall, while hypertension is relatively well-aligned, obesity and cholesterol in the synthetic dataset deviate significantly from the original, suggesting the need for better parameter tuning to improve realism.