---
title: "Data Analysis Exercise"
subtitle: "cdcdata-exercise"
author: "Murphy John"
format: html
editor: 
  markdown: 
    wrap: 72
---

Data Description: Data was downloaded from the CDC on 2025-02-03. This
dataset is titled NHANES Select Chronic Conditions Prevalence Estimates.
It contains national data on obesity, high total cholesterol, and
hypertension rates from 1999 to 2018 by sex, race, and age.There are a
total of 950 rows and 12 columns. See
[here](https://data.cdc.gov/NCHS/NHANES-Select-Chronic-Conditions-Prevalence-Estima/i667-sjhg/about_data)
for more details.

# Setup

```{r}
# load packages
library(dplyr)
library(ggplot2)
```

```{r}
# loads data
dat <- read.csv(here::here("cdcdata-exercise/data/raw/rawdata.csv"))
```

# Data processing

```{r}
# view data summary
summary(dat)

# select a subset of variables
dat1 <- dat %>%
  select(
    years = Survey.Years,
    sex = Sex,
    age = Age.Group,
    measure = Measure,
    percent = Percent
  ) 
```

```{r}
# filter for more granular data - seperate by sex, age categories
dat2 <- dat1 %>%
  filter(
    sex != "All",
    age != "20 and over",
    age != "2-19"
  ) %>%
  mutate(
    age_factor = factor(age, levels = c(
      "2-5","6-11","12-19","20-29","30-39","40-49",
      "50-59","60-69","70 and over")
    ),
    age = as.numeric(age_factor),
    sex = factor(sex, levels = c("Female", "Male")),
    years_factor = factor(years, levels = c(
      "1999-2000","2001-2002","2003-2004",
      "2005-2006","2007-2008","2009-2010",
      "2011-2012","2013-2014","2015-2016","2017-2018"
    )),
    years = as.numeric(years_factor)
  )

# turn obesity, cholesterol, and hypertension into their own variables
dat3 <- dat2 %>%
  group_by(years, sex, age) %>%
  tidyr::pivot_wider(
    names_from = measure,
    values_from = percent
  )
```

# Exploratory data analysis

```{r}
# number of rows
nrow(dat3)

# number of columns
ncol(dat3)

summary(dat3)

# print first few obs
head(dat3)

# print last few obs
tail(dat3)
```

It looks like there is some missing data. Locate those missing
observations.

```{r}
# count NAs by column
colSums(is.na(dat3))

# filter data for only rows with NA values
dat_na <- dat3 %>% filter(if_any(everything(), is.na))
```

`High Total Cholesterol` and `Hypertension` data was not collected for
those aged less than 19 years. Remove these observations to focus on the
relationship between the three measures.

# Briefly, data processing

```{r}
dat4 <- dat3 %>% 
  dplyr::rename(
    obesity = Obesity,
    hypertension = Hypertension,
    cholesterol = `High Total Cholesterol`
  ) %>%
  na.omit()

summary(dat4)
```

# More EDA

Explore each variable's categories and/or distributions.

```{r}
# categorical variables
table(dat4$years, dat4$sex)
table(dat4$years, dat4$age)
table(dat4$sex, dat4$age)

# numeric variables
summary(dat4$obesity)
ggplot(dat4, aes(x = obesity)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") 

summary(dat4$cholesterol)
ggplot(dat4, aes(x = cholesterol)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") 

summary(dat4$hypertension)
ggplot(dat4, aes(x = hypertension)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") 
```

Lets look at the relationship between years and our three measurements.

```{r}
# obesity
plot(dat4$years, dat4$obesity)
summary(lm(obesity ~ years, data=dat4))

# hypertension
plot(dat4$years, dat4$hypertension)
summary(lm(hypertension ~ years, data=dat4))

# cholesterol
plot(dat4$years, dat4$cholesterol)
summary(lm(cholesterol ~ years, data=dat4))
```

For every one unit increase in year, there is a $1.2$ unit increase in
obesity. For every one unit increase in year, there is a $0.9$ unit
decrease in cholesterol.

Lets look at the relationship between sex and our three measurements.

```{r}
# obesity
plot(dat4$sex, dat4$obesity)
summary(lm(obesity ~ sex, data=dat4))

# hypertension
plot(dat4$sex, dat4$hypertension)
summary(lm(hypertension ~ sex, data=dat4))

# cholesterol
plot(dat4$sex, dat4$cholesterol)
summary(lm(cholesterol ~ sex, data=dat4))
```

Male sex has $3.03$ unit decrease in obesity from female sex. Male sex
has $3.45$ unit decrease in cholesterol from female sex.

Lets look at the relationship between age and our three measurements.

```{r}
# obesity
plot(dat4$age, dat4$obesity)
dat_age1 <- dat4 %>%
  filter(
    age != 9
  )
summary(lm(obesity ~ age, data=dat_age1))

dat_age2 <- dat4 %>%
  filter(
    age %in% c(4, 9)
  )
summary(lm(obesity ~ age, data=dat_age2))


# hypertension
plot(dat4$age, dat4$hypertension)
summary(lm(hypertension ~ age, data=dat4))

# cholesterol
plot(dat4$age, dat4$cholesterol)
summary(lm(cholesterol ~ age, data=dat4))
```

Excluding the final age category, for every subsequent age group, there
is a $3.14$ unit increase in obesity. There is no difference in obesity
between groups 4 and 9. For every subsequent age group, there is a
$13.56$ unit increase in hypertension. For every subsequent age group,
there is a $1.64$ unit increase in cholesterol.

# EDA Conclusions

There are 10 2-year `years` categories from 1999-2018. For each `years`
category, there is an observation for each `sex` (Female, Male), and
within each sex, an observation for each of 6 age groups
(20-29,30-39,40-49,50-59,60-69,70 and over). This makes $6*2=12$
observations each year and $10*12=120$ total observations in the data.
Each observation has a value for percentage of population with
`obesity`, high `cholesterol`, and `hypertension`. 

Obesity overall has
mean 35.31 with (30.57, 40.90) quartiles. For every one unit increase in
year group, there is a $1.2$ unit increase in obesity. Male sex has
$3.03$ unit decrease in obesity from female sex. Excluding the final age
category, for every subsequent age group, there is a $3.14$ unit
increase in obesity. There is no difference in obesity between age
categories "20-29" and "70 and over".

Hypertension overall has mean 49.98 with (30.65, 70.90) quartiles. There
is no obvious relationship between hypertension and year or sex. For
every subsequent age group, there is a $13.56$ unit increase in
hypertension.

Cholesterol overall has mean 14.35 with (8.5, 18.55) quartiles. For
every one unit increase in year, there is a $0.9$ unit decrease in
cholesterol. Male sex has $3.45$ unit decrease in cholesterol from
female sex. For every subsequent age group, there is a $1.64$ unit
increase in cholesterol.

# END OF PART 1
